---
title: "Katase et al 2026_KaplanMeierPlots_6a_6b"
output: html_document
---



In this R Markdown document, the Kaplan Meier curves will be reproduced from figures 6a and 6b from Katase et al 2026.
```{r}


library(tidyverse)
library(ggpubr)
library(survival)
library(survminer)
library(TCGAbiolinks)
library(SummarizedExperiment)
library(annotables)
library(here)
library(patchwork)


#Pull Gene Names from BioMRT

Gene_Symbols1 <- grch38 %>% dplyr::select(ensgene, symbol)

Figure_Export <- file.path(here(),"Figures")

Data_Import <- file.path(here(), "Data")

```


```{r}


clinical_data <- GDCquery_clinic("TCGA-HNSC", type = "clinical", save.csv = FALSE)


clinical_data <- clinical_data %>% dplyr::select(race, gender, vital_status, age_at_index, cause_of_death, days_to_diagnosis, days_to_recurrence, days_to_death, days_to_last_follow_up, treatments_pharmaceutical_treatment_intent_type, treatments_radiation_treatment_outcome, classification_of_tumor, ajcc_clinical_stage, bcr_patient_barcode)


###

#These have been previously downloaded and extracted from the TCGA Biolinks package and then will be imported faster though this reading an RDS file. If you need to import it directly again from the website or through the package, check one of the DEG_Fig1 file where I perform this.

hnsc_metadata <- readRDS(file.path(Data_Import, "hnsc_metadata.rds"))


hnsc_coldata <- readRDS(file.path(Data_Import, "hnsc_coldata.rds"))


hnsc_coldata_specified <- hnsc_coldata %>% dplyr::select(patient, sample_type, sample) %>% rownames_to_column(., var = "full_patient_ID")

```

#Convert Clinical Data Variables into a status that I can use in survival analyses

The survival package requires that the vital_status be in binary, or 0/1, depending on whether one is alive or dead respectively. Therefore, a new variable is created indicating whether the patient is alive or dead and how long they survived before they left the study (either due to being censored, or dying).

```{r}
clinical_data <- clinical_data %>%
      mutate(survival_length = if_else(vital_status == "Alive", days_to_last_follow_up, days_to_death)) #This line of code makes a total survival length plot where if they are still alive, it shows the length as the "days to last followup" where if they died, it shows days_to_death.
```


#Import TPM Count

Although Katase et al 2026 seems to indicate that they used the TPM_unstranded_counts which I am importing saved data that was previously downloaded from the TCGA database. The TPM raw unstranded counts were imported and then manipulated to put them into a proper format for later graphing and survival analysis. Another alternative would be to use the VST stabilized counts calculated by the DESeq() function, but that is not what was performed by Katase et al 2026. When there were duplicate ensgene IDs, which is likely because of the different gene transcript for each gene, I just added the sum of the expression together and combined them. Considering the duplicates all had values of 0, it should not change the results.

```{r}


transformed_data <- readRDS(file.path(Data_Import,"tpm_unstranded_raw.rds"))

filtered2 <- transformed_data %>%
      as.data.frame() %>%
      rownames_to_column(., var = "ensgene") %>%
      mutate(ensgene = str_remove(ensgene,"\\..*")) %>% group_by(ensgene) %>% summarize(across(everything(), sum)) #Needed because of duplicates in the ENS gene ID (likely because of different transcripts)

rm(transformed_data)

filtered3 <- filtered2 %>% 
      left_join(., Gene_Symbols1, by = "ensgene") %>%
      dplyr::select(-ensgene) %>%
      pivot_longer(., cols = -(symbol)) %>%
      mutate(name = str_replace(name, "-01A.*", "-01A"), name = str_replace(name, "-11A.*","-11A"), name = str_replace(name, "-01B.*","-01B")) #Remove the extra values from name that are not needed extra values so that data frames can be joined together in the next step. -01A is correlated to a tumor sample, and -11A is correlated with a normal. tissue sample.
      
#The High or Low values were calculated by each individual sample was over 50% mean expression of that gene (only in "Primary Tumor" tissue). Since they did not use the Solid Normal Tissue with respect to survival analysis, they were removed prior to this step.

filtered <- filtered3 %>%
      group_by(symbol) %>% 
      left_join(., hnsc_coldata_specified, by = c("name" = "sample")) %>%
      filter(sample_type == "Primary Tumor") %>%
      mutate(Expression_Level = if_else(value > mean(value), "High", "Low")) %>%
      ungroup()

ClinicalData_and_Counts <- filtered %>%
      left_join(., clinical_data, by = c("patient" = "bcr_patient_barcode")) %>%
      mutate(deceased = if_else(vital_status == "Alive", 0, 1),
             time_in_months = survival_length / 30.41)

#Save the Clincal Data with Counts as an object for later R analysis
#saveRDS(object = ClinicalData_and_Counts, file = file.path(Data_Import, "Clinical_Data_and_Final_Count.rds"))
```


The next step is to filter for the top 10 genes that were upregulated from the Differential Gene Expression.

```{r}

upregulated_order <- c("CA9","MMP11","IL11","COLGALT1","COL10A1", "C5orf34-AS1", "FOXD2-AS1", "HOXC6", "CCN4","CDCA5") 

upregulated_vst <- filtered %>%
      filter(symbol %in% upregulated_order)

```


#Kaplan Survival Curve for the CA9 gene

In this code chunk, a survival curve will be generated for
```{r}
GENE <- "CA9"

final_data <- upregulated_vst %>%
    #  filter(symbol == GENE) %>%
      left_join(., clinical_data, by = c("patient" = "bcr_patient_barcode")) %>%
      mutate(deceased = if_else(vital_status == "Alive", 0, 1),
             time_in_months = survival_length / 30.41) 

fit <- survfit(Surv(time_in_months, deceased) ~ Expression_Level, data = final_data)

plot <- ggsurvplot(fit,
           data = final_data,
           pval = T, 
           legend = "right",
           title = GENE,
           fun = "pct")

Strata_count <- plot$data.survtable %>% filter(pct.risk == 100)
High_N <- Strata_count[1,"n.risk"]
Low_N <- Strata_count[2,"n.risk"]

plot$plot + theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      theme(legend.position = c(0.8, 0.8))+ labs(x = "months")+
      scale_color_manual(values = c("Expression_Level=Low" = "black", "Expression_Level=High" = "red"),
                         labels = c(paste0("High (N=",High_N,")"), paste0("Low (N=",Low_N,")"))) #could use glue or paste

```

Survival Graphs Function

```{r}

Survival_Graphs <- function(data = data_matrix, 
                             order = NA, newcol = 5, 
                             newrow = 2, total_graphs = 10,
                             gene_identifier = "symbol",
                             time = "survival_length",
                             status = "deceased",
                             measureable_metric = "Expression_Level",
                             x_axis = "Time",
                             y_axis = NULL){
      
#Filters what column is chosen to be the unique gene identifier      
      Symbols <- unique(as.character(data[[gene_identifier]]))

 
#If no order is given to make the graphs, the the symbol order stays whatever it is above           
      if (is.na(order[1]) == FALSE){
            
       Symbols <- order
       
      }
      

#Create a vector for up to 10 empty plots
     FINAL_PLOT <-  vector("list", 10)
      
 #Run a loop for each Gene Symbol/ensgene
     
      for (i in 1:length(Symbols)){
      
      Symbol_now <- Symbols[i]

#Select the rows that have the specific gene I am looking for:
      
slice_rows <- grepl(pattern = Symbol_now, x = c(data[[gene_identifier]]))


survival_curve <- data[slice_rows,]

#Convert data to a dataframe so it works with ggsurvplot. 
survival_curve <- as.data.frame(survival_curve)

#Extract the Title equal to the gene_name

GENE_NOW <- Symbol_now


#Extract the outcome timeline, alive or dead (survival) metric, as well as the measurable metric

time1  <- as.numeric(survival_curve[[time]])
status1 <- as.numeric(survival_curve[[status]])
group1  <- as.factor(survival_curve[[measureable_metric]])

survival_curve2 <- data.frame(time = time1, status = status1, group = group1)



#Run Survival analysis. I had to do it by making a formula and then importing it into the ggsurvplot for some reason with the internal structure that I had to use some AI to help me figure out, but this worked.

#KAPLAN <- survfit(as.formula("Surv(time, status) ~ group"), data = survival_curve2)
KAPLAN <- survfit(Surv(time, status) ~ group, data = survival_curve2)
#print(KAPLAN)

g <- ggsurvplot(fit = KAPLAN, data = survival_curve2, pval = TRUE, legend = "right", title = GENE_NOW, fun = "pct") 


strata_count <- g$data.survtable


Strata_count <- strata_count %>% filter(pct.risk == 100)
High_N <- Strata_count[1,"n.risk"]
Low_N <- Strata_count[2,"n.risk"]

#Modulate the title of the plot
FINAL_PLOT[[i]] <- g$plot + theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      theme(legend.position = c(0.75, 0.75)) + labs(x = x_axis)+
      scale_color_manual(values = c("group=High" = "red", "group=Low" = "black"),
      labels = c(paste0("High (N=",High_N,")"), paste0("Low (N=",Low_N,")"))) #Change the color


      }
       
     
     #Combine all the plots together 
     FINAL_PLOT <- wrap_plots(FINAL_PLOT, ncol = newcol)
      
      print(FINAL_PLOT)    
}



```


#Top 10 Upregulated Genes Survival Curves

```{r}
png(filename = file.path(Figure_Export,"Top_10_upregulated_Genes_KaplanMeier_Fig6a.png"),
    width = 2050, height = 1000, units = "px")


order <- c("CA9","MMP11","IL11","COLGALT1","COL10A1", "C5orf34-AS1", "FOXD2-AS1", "HOXC6", "CCN4","CDCA5") 

final_data %>%
      Survival_Graphs(data = ., order = order, x_axis = "OS Months", time = "time_in_months")

dev.off()

```


#Top 10 Downregulated Survival Curves
```{r}
downregulated_order <- c("LINC01829","NUCB2","AGFG2","IL12A-AS1","CAB39L", "CCL28","NOSTRIN","GPD1L","EMP1","SLC8A1-AS1") 

downregulated_set <- filtered %>%
      filter(symbol %in% downregulated_order)

downplot <- downregulated_set %>%
      left_join(., clinical_data, by = c("patient" = "bcr_patient_barcode")) %>%
      mutate(deceased = if_else(vital_status == "Alive", 0, 1),
      time_in_months = survival_length / 30.41) %>%
      Survival_Graphs(data=., order = downregulated_order, x_axis = "OS Months", time = "time_in_months")

png(filename = file.path(Figure_Export, "Top_10_downregulated_Genes_KaplanMeier_Fig6b.png"),
    width = 2050, height = 1000, units = "px")

downplot

dev.off()

```




#Survival Graph group_by


Excess and unneeded code where I was attempting to use the ggsurvplot_facet() and group_by() functions to do what I did in my above function faster. However, the result were unable to be formatted in the way that was desired.
```{r}
data1 <- upregulated_vst %>%
      #filter(symbol == GENE) %>%
      left_join(., clinical_data, by = c("patient" = "bcr_patient_barcode")) %>%
      mutate(deceased = if_else(vital_status == "Alive", 0, 1),
             time_in_months = survival_length / 30.41) %>% as.data.frame()
    
fit <- survfit(Surv(survival_length, deceased) ~ Expression_Level, data = data1)

ggsurvplot_facet(fit = fit, data = data1, facet.by = "symbol",
           pval = TRUE, 
           legend = "right",
           fun = "pct")

ggsurvplot_group_by(fit = fit, data = data1, group.by = "symbol",
           pval = TRUE, 
           legend = "right",
           fun = "pct")


```



