---
title: "Katase_et_al_2026_PCA"
output: html_document
---


#Step 1): Activate the proper libraries

```{r}
library(tidyverse)
library(TCGAbiolinks)
library(survminer)
library(survival)
library(SummarizedExperiment)
library(DESeq2)
library(annotables)
library(here)

Figure_Export <- file.path(here(),"Figures")

Data_Import <- file.path(here(), "Data")


```

#Step 2): Import the TCGA biolinks data (and we aren't going to be doing any data analysis where we only pick maybe 20 samples of each!)

In this case, I will be downloading transcriptome profiling data from the TCGA database because I want to be measure the transcriptome to compare expression of different genes in TCGA-HNSC dataset to compare.
```{r}
file1 <- file.path(Data_Import, "vst_transformed_data.rds")

vst_counts <- readRDS(file1)

Gene_Symbols1 <- grch38 %>% select(ensgene, symbol) %>% distinct()

#Gene Names
Gene_Symbols <- grch38 %>% dplyr::select(ensgene, symbol)
```


#Step 3: Import the clinical metadata
```{r}
clinical_data <- GDCquery_clinic("TCGA-HNSC",
                                 type = "clinical", save.csv = FALSE)
```

#Step 4: Download Data from the new query so that I can extract the counts if needed
```{r cache= TRUE}
GDCdownload(TCGA_HNSC)
```

#Step 5: Extract the counts from the dataset
```{r}
hnsc_data <- GDCprepare(TCGA_HNSC, summarizedExperiment = TRUE) #Extract the data using GDCprepare
hnsc_matrix <- assay(hnsc_data, "tpm_unstrand") #I can use the assay() function to get the tpm results?
```


#Step 6: Get the raw_data and metadata and column data

```{r}
#Get Metadata and ColData

hnsc_coldata <- readRDS(file = file.path(Data_Import,"hnsc_coldata.rds"))

hnsc_metadata <- readRDS(file = file.path(Data_Import,"hnsc_metadata.rds"))
```


#Lets do some PCA analysis using the vst data (which I know is not as good and I should use the normalized counts, but I haven't figured that out yet)

In this example, the biggest problem I am having is the fact that multiple ENS gnee IDs map to the same gene (or multiple different transcripts map to the same gene) and consequently there is a duplicate of the symbols which is causing issues when I try to run the analysis because there are "duplicates" of certain genes. What I am going to do to account for that is just


https://www.researchgate.net/post/How-to-deal-with-multiple-ensemble-IDs-mapping-to-one-gene-symbol-in-a-RNA-Seq-dataset

```{r}
library(MASS)
library(factoextra)

#Orient the raw_count data into the proper form so that it can be transposed later. In this example, the biggest problem I am having is that
hnsc_pca <- vst_counts %>%
  as.data.frame() %>%
  rownames_to_column(., var = "ensgene") %>% #It appears mutate(ensgene = str_remove(ensgene,"\\..*")) %>% #This is the regular expression to remove everything after the . in ensgene, I will admit I don't fully understand it yet and had to use AI to help on this.
  group_by(ensgene) %>%
  summarize(across(where(is.numeric), sum)) %>%
  ungroup() %>%
  left_join(.,Gene_Symbols1, by = "ensgene") %>%
  distinct() # I am not sure why, but it duplicates a lot of rows with the same values, therefore I can just remove the non-unique rows. Some of the code I used to figure that out is below, but not in a code chunk. I suspect if I was doing actual analysis for a paper, I might have to look more closely. However, when it functions, there seems to be clear duplication for each column which is likely affecting analysis.

  
# meta_data simplified which will be needed for the matrix
pca_meta <- hnsc_coldata %>%
  remove_rownames() %>%
  dplyr::select(barcode, sample_type, tumor_descriptor, ajcc_pathologic_stage, primary_diagnosis, prior_malignancy, classification_of_tumor)

#Make the transposed matrix to input into the PCA
hnsc_pca_transposed <- hnsc_pca %>%
  dplyr::select(-symbol) %>%
  t() 


hnsc_pca_transposed <- hnsc_pca_transposed %>% 
as.data.frame() %>%
mutate(across(everything(), as.numeric)) %>%
rownames_to_column(., var = "barcode") %>%
left_join(., pca_meta, by = "barcode") 

#Remove the excess row
hnsc_pca_transposed <- hnsc_pca_transposed[-c(1),]


  #Converting the transposed_dataset into the data frame / matrix that will be input into the PRCOMP function to make the PCA table
input_matrix <- hnsc_pca_transposed %>%
  dplyr::select(-c(barcode, sample_type, tumor_descriptor, ajcc_pathologic_stage, primary_diagnosis,prior_malignancy,classification_of_tumor))


colnames(input_matrix) <- hnsc_pca$ensgene 

PCA_result <- prcomp(input_matrix, 
                     scale = FALSE) #Preferred but I can't do it in this case

fviz_pca_biplot(PCA_result,
               label = "none",
                habillage = hnsc_pca_transposed$sample_type,
                geom.var = c(""))
  
```

## Linear Discriminant Analysis

https://www.youtube.com/watch?v=azXCzI57Yfc


```{r}
PCA_Matrix <- as.data.frame(PCA_result$x)

#Ensure that the matrix variable (which differentiates the different grouops) is accurate to what we are trying to separate
PCA_Matrix$sample_type <- as.factor(hnsc_pca_transposed$sample_type)

#Run the LDA model

LDA_Model <- lda(sample_type ~ PC1 + PC2 + PC3, data = PCA_Matrix)


#Make a prediction which I can then use to graph the data
lda_predict <- predict(LDA_Model)

#Make a dataframe with the predictions using the predict function. Based on what copilot said, there is only one LD factor because it is equal to the number of different observations - 1. So if I have tumor and normal tissue, that is 2 unique factors - 1 which is equal to 1. This is different to the analysis before which had 4 different factors.
lda_df <- data.frame(
  LD1 = lda_predict$x[,1],
  LD2 = lda_predict$x[,2],
  Sample_Type = PCA_Matrix$sample_type
)

#Make a graph (Since I only have 1 discriminant factor in this example, it looks like the normal LDA graph isn't very good)
lda_df %>%
  ggplot(aes(x=LD1, fill = Sample_Type))+
  geom_density(alpha = 0.2 )+
  theme_classic()

```


```{r}

dds <- readRDS(file = file.path(Data_Import, "dds_object.rds"))

vst_dds <- vt(dds, blind = TRUE)

PCA_plot <- plotPCA(vst_dds, ntop = 60000, intgroup = "sample_type")

```

