---
title: "Survival_Analysis_CDK1_Fig9"
output: html_document
---

#Import relevant libraries

```{r}
library(tidyverse)
library(here)
library(SummarizedExperiment)
library(survminer)
library(survival)
library(TCGAbiolinks)

Figure_Export <- file.path(here(),"Figures")

Data_Import <- file.path(here(), "Data")
```


#Import the TCGA Counts

Import the CPM (Counts per Million) from the TCGA-LIHC Dataset that were previously generated. Additionally, the Log2(CPM + 1) counts were imported to see if that causes any difference in the results. It turns out that it doesn't make a difference which I didn't expect, but I wanted to test it for sure.
```{r}
#Extract Counts
TCGA_CPM_Counts <- readRDS(file.path(Data_Import, "CPM_counts.rds"))

TCGA_Log2_CPM_Count <- as.data.frame(apply(TCGA_CPM_Counts, 2, FUN = function(x){log2(x + 1)}))


#Extract Coldata
TCGA_coldata <- readRDS(file.path(Data_Import, "LIHC_coldata.rds"))

TCGA_coldata <- TCGA_coldata %>%
  dplyr::select(barcode, definition, patient)

#Import the Clinical Data

TCGA_LIHC_Clinical_Data <- GDCquery_clinic("TCGA-LIHC",
                                 type = "clinical", save.csv = FALSE)

```

#Initial Data Formating

This chunk will have the count matrices into dataframes and then filtering by the gene of interest (CDK1). Then, the data is pivoted into a form where the coldata can be merged and the data be filtered to remove any patient samples that were non-tumors. This is because we do not need the 5 patients who gave non-tumor tissue but did not have any cancer. Furthermore, we need to remove the samples 
```{r}

TCGA_CDK1 <- TCGA_CPM_Counts %>%
  as.data.frame() %>%
  rownames_to_column(., var = "Symbol") %>%
  filter(Symbol == "CDK1") %>%
  pivot_longer(data=., cols = -c("Symbol"), names_to = "barcode", values_to = "Counts") %>%
  left_join(., TCGA_coldata, by = "barcode") %>%
  filter(definition == "Primary solid Tumor")


TCGA_CDK1_LOG2 <- TCGA_Log2_CPM_Count %>%
  as.data.frame() %>%
  rownames_to_column(., var = "Symbol") %>%
  filter(Symbol == "CDK1") %>%
  pivot_longer(data=., cols = -c("Symbol"), names_to = "barcode", values_to = "Counts") %>%
  left_join(., TCGA_coldata, by = "barcode") %>%
  filter(definition == "Primary solid Tumor")
```

#Format Clinical_Data

The Clinical Data needs to be formatted to ensure that the alive/dead status is numeric where 1 means they are dead and 0 means they are alive (at the end of the time where they "leave" the study). If they are alive, it means they were censored and left for another reason. If they died, it counts as a death for the survival analysis. Furthermore, the days of survival is converted into months to be consistent with the graphs from Zhu et al 2018.

```{r}
TCGA_LIHC_Clinical_Data <- TCGA_LIHC_Clinical_Data %>%
  mutate(Vital_Status_Numeric = if_else(vital_status == "Alive", 0, 1)) %>% #A death is considered 1 and being alive is 0 (or censored if they were removed before the end of the timeline)
  mutate(Overall_Survival = if_else(vital_status == "Dead",days_to_death, days_to_last_follow_up),
         Overall_Survival = as.numeric(Overall_Survival),
         Overall_Surv_months = Overall_Survival / 30.417) #Conversion of Survival Days into Months

#Combine the Clinical Data with Count Data 

#Combined with the CPM normalized data
TCGA_CDK1 <- TCGA_CDK1 %>%
  left_join(., TCGA_LIHC_Clinical_Data, by = c("patient" = "submitter_id")) %>% 
  dplyr::select(barcode, patient, Counts, definition, days_to_death, days_to_diagnosis, days_to_last_follow_up, vital_status, Vital_Status_Numeric,
                Overall_Survival, Overall_Surv_months) %>%
  mutate(Expression_Level = if_else(Counts > median(Counts), "High", "Low")) #Convert each patient sample into high or low expression
  

#Combined with the log2(CPM + 1) Data
TCGA_CDK1_LOG2 <- TCGA_CDK1_LOG2 %>%
  left_join(., TCGA_LIHC_Clinical_Data, by = c("patient" = "submitter_id")) %>% #Convert each patient sample into high or low expression
  dplyr::select(barcode, patient, Counts,  days_to_death, days_to_diagnosis, days_to_last_follow_up, vital_status, Vital_Status_Numeric,
                Overall_Survival, Overall_Surv_months) %>%
   mutate(Expression_Level = if_else(Counts > median(Counts), "High", "Low")) 
```


#Kaplan Meier Plot for the TCGA-LIHC Dataset with the CPM counts from the matrix

```{r}
#Run Survival Model
TCGA_SURVIVAL_MODEL <- survfit(Surv(Overall_Surv_months, Vital_Status_Numeric) ~ Expression_Level, data = TCGA_CDK1)

#Extract Plot
TCGA_plot <- ggsurvplot(fit = TCGA_SURVIVAL_MODEL,
           pval = T,
           surv.scale = "percent"
           )
#Make Cosmetic Changes to plot
a <- TCGA_plot$plot +
  labs(x = "Time (months)",
       y = "Overall Survival (%)")+
  theme(legend.position = "right")+
  scale_color_manual(values = c("Expression_Level=High" = "red", "Expression_Level=Low" = "blue"),
                     labels = c("Expression_Level=High" = "High", "Expression_Level=Low" = "Low"))

#Graph Plot
a

#Save plot
ggsave(file = "TCGA-LIHC_CDK1_Kaplan_Meier_10a.png",
       path = Figure_Export)
```

#Kaplan Meier Plot for the TCGA-LIHC Dataset with the Log2(CPM + 1) counts from the matrix

```{r}
TCGA_SURVIVAL_MODEL_LOG2 <- survfit(Surv(Overall_Surv_months, Vital_Status_Numeric) ~ Expression_Level, data = TCGA_CDK1_LOG2)

ggsurvplot(fit = TCGA_SURVIVAL_MODEL_LOG2,
           pval = T,
           surv.scale = "percent"
           )
```



## Repeat with the GEO Dataset

```{r}
#Import of Microarray and Clinical Data
Normalized_Microarray <- readRDS(file = file.path(Data_Import, "Normalized_Microarray.rds"))

Microarray_Clinical_data <- readRDS(file = file.path(Data_Import, "Microarray_Clinical_data.rds"))

#Import of Probes for the Microarray
library(GEOquery)
GSE <- getGEO(
  GEO = "GSE14520",
  GSEMatrix = TRUE
)

#Extract the Probe IDs for each of the Probes (For the GPL3921 and GPL571 respectively)
probe3921 <- GSE$`GSE14520-GPL3921_series_matrix.txt.gz`@featureData@data
probe3921 <- probe3921 %>% dplyr::select(ID, GPL3921_GENESYMBOL = `Gene Symbol`)

probes571 <- GSE$`GSE14520-GPL571_series_matrix.txt.gz`@featureData@data
probes571 <- probes571 %>% dplyr::select(ID, GPL571_GENESYMBOL = `Gene Symbol`)
```

#Formatting the Normalized_Microarray Data and Clinical Data

```{r}
#Combining the probes with the microarray data
Normalized_Microarray <- Normalized_Microarray %>%
  rownames_to_column(., var = "ID") %>%
  left_join(., probe3921, by = "ID") %>% #GPL3921 probes
  left_join(., probes571, by = "ID")     #GPL571 probes

#Formatting the Microarray data which includes the last 3 steps which sum up the counts for when multiple probes are correlated to the same gene. There is a chance that the probes are double counting, or perhaps they target different transcripts or isoforms that all should be counted. Those steps can be removed if needed in the future

Normalized_Microarray <- Normalized_Microarray %>%
  pivot_longer(data=., cols = -c("GPL571_GENESYMBOL", "GPL3921_GENESYMBOL", "ID"), names_to = "geo_accession", values_to = "Count") %>%
  mutate(Symbol = paste(GPL3921_GENESYMBOL, "or", GPL571_GENESYMBOL)) %>% 
  dplyr::select(-c("GPL3921_GENESYMBOL", "GPL571_GENESYMBOL", "ID")) %>%
  filter(Symbol == "CDC2 or CDK1") %>% #Filter or the specific gene of interest ******GENE OF INTEREST********
  group_by(geo_accession, Symbol) %>% 
  summarize(Count = sum(Count)) %>% 
  ungroup()
  
```

#Combining Microarray Data with Microarray Clinical Data

```{r}
Normalized_Microarray <- Normalized_Microarray %>%
  left_join(., Microarray_Clinical_data, by = c("geo_accession" = "Affy_GSM")) %>%
  filter(Tissue.Type == "Tumor") %>%
  mutate(Expression_Level = if_else(Count > median(Count), "High", "Low"))
```

#Generate Kaplan-Meier Plot from Microarray Data

```{r}
#Run Model
GEO_surv <- survfit(Surv(Survival.months, Survival.status) ~ Expression_Level, data = Normalized_Microarray)

#Extract Plot
GEO_plot <- ggsurvplot(fit = GEO_surv,
           pval = T,
           surv.scale = "percent"
           )

#Make Cosmetic Changes
p <- GEO_plot$plot +
  labs(x = "Time (months)",
       y = "Overall Survival (%)",
       title = "Kaplan-Meier plot of TCGA-LIHC Dataset on CDK1 Expression")+
  theme(legend.position = "right",
        plot.title = element_text(face = "bold", hjust = 0.5))+
  scale_color_manual(values = c("Expression_Level=High" = "red", "Expression_Level=Low" = "blue"),
                     labels = c("Expression_Level=High" = "High", "Expression_Level=Low" = "Low"))

#Graph Plot 
p 

#Save
ggsave(file = "GEO_GSE14520_CDK1_Kaplan_Meier_10b.png",
       path = Figure_Export)
```

