---
title: "Dawadi_et_al_2025"
output: html_document
---

Link to github with "data" that I will be using: https://github.com/dprabin25/BegineersRNA-Seq

#Paper that discusses RNA-SEQ practices for begginers
https://pmc.ncbi.nlm.nih.gov/articles/PMC12599993/

PCA Pipeline for Genomic Data

1) Import Data
2) Make sure sample IDs in metadata match column names of the raw_counts
3) For the DeSEQ2 function, make sure the meta_data rownames are equal to the sample IDs which should be *identical* to the column names of the raw counts
4) Deal with any duplicates in ENS IDs or Gene Names
5) Run the DeSEQ2 Analysis (it needs raw counts not normalized counts because it automatically normalizes the counts for us)
6) Optional? --> Remove any GeneIDs where the rowSums of the raw counts is very small to speed up the analysis. If the raw counts # is very low, its unlikely to have any major effect on the differential expression analysis
7) Run the vst() function on the DESeq object
8) Make a PCA Plot using the plotPCA function from dplyr OR use the library(MASS) and use the other PCA software



#Step 1: Activate Appropriate Libraries

I include many, but not all, of the libraries here
```{r}
library(tidyverse)
library(TCGAbiolinks)
library(survminer)
library(survival)
library(SummarizedExperiment)
library(DESeq2)
library(annotables)
library(here)
```

#Step 2: Import the data
```{r}
file <- file.path(here(), "/GEO_Database/Dawadi_et_al_2025/Data/raw_counts.csv")

Dawadi_raw_counts <- read.csv(file =file)

meta_data_dawadi <- read.csv(file.path(here(),"/GEO_Database/Dawadi_et_al_2025/Data/MetaData.csv"))

meta_data_dawadi <- meta_data_dawadi %>%
 dplyr::rename(Biopsy_Site = `Sample.Characteristic.biopsy.site.`)

GeneNames <- read_tsv(gzcon(url("https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts&file=Human.GRCh38.p13.annot.tsv.gz")))

Gene_Names <- GeneNames %>% dplyr:: select(GeneID, Symbol, EnsemblGeneID)


```

#Step 3: Check to make sure the Sample IDs in the metadata matches the Sample IDs in the raw_count data
```{r}
#Check to make sure the samples IDs in the metadata matches the raw_count

Dawadi_raw_counts_2 <- Dawadi_raw_counts %>% column_to_rownames(., var = "Gene.ID") %>% dplyr::select(-Gene.Name)

identical(meta_data_dawadi$Run, colnames(Dawadi_raw_counts_2))
          
```

#Step 4: Look for duplicates in the raw_counts

Look for any duplicates in the raw_counts and decide what I want to do with them. One option is to just add them all together since its likely slightly different transcripts for the same gene. However, it depends on what one wants to look at.

```{r}
sum(duplicated( Dawadi_raw_counts_2$Gene.ID))

# There are no duplications which makes this easy


```

#Step 5: Run the DESeq2 Analysis

```{r}
DESEQ <- DESeqDataSetFromMatrix(countData = Dawadi_raw_counts_2,
                                colData = meta_data_dawadi,
                                design = ~ Biopsy_Site)

#Create the vst transformed data for later analysis
vst_dawadi <- vst(DESEQ, blind = TRUE)
vst_dawadi_dataframe <- as.data.frame(assay(vst_dawadi))

vst_dawadi_dataframe <- vst_dawadi_dataframe 
  
write.csv(vst_dawadi_dataframe, file = file.path(here(),"/GEO_Database/Dawadi_et_al_2025/Data/vst_transformed_counts.csv"))
```


#Step 6: Plot PCA by using the plotPCA function from DESeq()
```{r}
png(filename = file.path(here(),"/GEO_Database/Dawadi_et_al_2025/plotPCA_Fig5E.png"),
  width = 480, height = 480, units = "px", pointsize = 12)

plotPCA(vst_dawadi, intgroup = "Biopsy_Site", ntop = 1000)

dev.off()


```

#Step 7: Plot PCA using the other software via manual manipulation and the factoextra package()

It should be noted that in this example, we are using all the genes as opposed to the top 1000 genes expressed like in Step 6. That is likely one of the contribution factors why PC1 and PC2 involved in this graph don't have as high of a percentage of explaining the variation as the other graph.
```{r}
library(factoextra)
library(MASS)

PCA_full <-  vst_dawadi_dataframe %>%
  t() %>% 
  as.data.frame() %>%
  rownames_to_column(., var = "Run") %>%
  left_join(.,
            meta_data_dawadi %>% dplyr::select(Run, Biopsy_Site), 
            by = "Run")

PCA_input <- PCA_full %>% dplyr::select(-Run, -Biopsy_Site)


PCA_results <- prcomp(PCA_input, scale = FALSE)

png(filename = file.path(here(),"/GEO_Database/Dawadi_et_al_2025/fviz_pca_biplot_Fig5E.png"),
  width = 480, height = 480, units = "px", pointsize = 12)

fviz_pca_biplot(PCA_results,
                label = "none",
                habillage = PCA_full$Biopsy_Site)


dev.off()


```

