---
title: "Dawadi_et_al_2025"
output: html_document
---

Link to github with "data" that I will be using: https://github.com/dprabin25/BegineersRNA-Seq

#Paper that discusses RNA-SEQ practices for begginers
https://pmc.ncbi.nlm.nih.gov/articles/PMC12599993/

To be clear, this is a very small dataset but I want to be able to reproduce everything here and eventually move on to better and more detailed files/analysis.

General Pipeline:
) Import Data
2) Format the Meta Data
3) Format the Raw Count Data and look for potential duplicates to account for 
4) For the DeSEQ2 function, make sure the meta_data rownames are equal to the sample IDs which should be *identical* to the column names of the raw counts
5) Run the DeSEQ2 Analysis (it needs raw counts not normalized counts because it automatically normalizes the counts for us)
6) Optional? --> Remove any GeneIDs where the rowSums of the raw counts is very small to speed up the analysis. If the raw counts # is very low, its unlikely to have any major effect on the differential expression analysis
7) Convert the DeSEQ2 analysis into a results object 
8) Convert the results object into a dataframe
9) Make graphs as wanted/needed


#Step 1: Activate Appropriate Libraries

I include many, but not all, of the libraries here
```{r}
library(tidyverse)
library(TCGAbiolinks)
library(survminer)
library(survival)
library(SummarizedExperiment)
library(DESeq2)
library(annotables)
library(here)
```

#Step 2: Import the data
```{r}
file <- file.path(here(), "/GEO_Database/Dawadi_et_al_2025/Data/vst_transformed_counts.csv")

transformed_count <<- read.csv(file = file)

meta_data_dawadi <- read.csv(file.path(here(),"GEO_Database/Dawadi_et_al_2025/Data/MetaData.csv")) 

meta_data_dawadi <- meta_data_dawadi %>%
 dplyr::rename(Biopsy_Site = `Sample.Characteristic.biopsy.site.`)


```

#Step 3:

In this simple example, to make it easier, I am going to randomly sample genes with 10 separate ENS Ensemble IDs and use them in the heatmap. I had to manually extract the specific ENS Ensemble IDs that they used by hand so I could reproduce the graph. The resultant graph is similar, but not exactly the same because I am not going to bother spending lots of time changing the other cosmetic factors to make the graph look like the one they gave. However, the data is accurate.
```{r}
ENS_id_endings <- c("113742","240498","172159","204936","071203","148180","182271","174992","123028","16602","168079","157434","165816","101187","011412",
                    "15413", "215030", "130649", "161944","257017")

for (i in 1:length(ENS_id_endings)) {
  ENS_ID[i] <- paste0("ENSG00000", ENS_id_endings[i])
} 

transformed_count2 <- transformed_count %>%
  dplyr::select(-X) %>%
  filter(EnsemblGeneID %in% ENS_ID) %>%
  column_to_rownames(., var = "EnsemblGeneID") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(., var = "Run") %>%
  left_join(., meta_data_dawadi %>% dplyr::select(Run ,Biopsy_Site), by = "Run") %>%
  group_by(Biopsy_Site) %>%
  mutate(Biopsy_Site = paste0(Biopsy_Site,"_",as.character(row_number()))) %>% 
  dplyr::select(-Run) %>%
  column_to_rownames(., var = "Biopsy_Site")

transformed_count2 <- as.matrix(transformed_count2) 


heatmap(transformed_count)

library(pheatmap)
pheatmap(transformed_count2)


```

